Linear analysis of the vertical shear instability in astrophysical disks

### 2015-01-27 application.tex
completed rough summary

\subsection{Newtonian cooling}
For $L\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

@@

\subsection{Toy model for realistic thermal relaxation}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff} =
  \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
and Eq. \ref{eta_def}. Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$.

@@

\subsection{Numerical example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The disk is unstable to the fundamental isothermal VSI, but not
strongly.

>>>>

\subsection{Newtonian cooling}\label{newton_cool}
For $L\ll l_\mathrm{rad}$ the radiative cooling is given by {\bf (reference?)}
\begin{align}
\nabla\cdot\bm{F} = 4 \rho \kappa_d \sigma_s T^4,
\end{align}   
where $\sigma_s$ is the Stefan-Boltzmann constant. We choose the
heating term 
\begin{align}
  Q_+ = 4\rho\kappa_d\sigma_s T^{4}_{t=0}. 
\end{align}
Then Eq. \ref{real_energy} becomes
\begin{align}
  \frac{DP}{Dt} = -\gamma P \nabla\cdot\bm{v} -
  \frac{4P\kappa_d\sigma_s}{Tc_v}\left(T^4 - T_{t=0}^4\right). 
\end{align}
When linearized, the last term on the RHS becomes
\begin{align}
  -\frac{16\sigma_s}{c_v}\kappa_d T^3\left(\delta P -
    \frac{P}{\rho}\delta\rho\right). 
\end{align}
By comparing this expression with the linearized version of the energy
equation used in our calculations (Eq. \ref{energy_eq}), we identify 
\begin{align}\label{tc_newton_cool} 
  t_\mathrm{cool} = \frac{c_v}{16\sigma_s\kappa_dT^3}
\end{align}
as the thermal relaxation timescale for perturbation lengthscales
$L\ll l_\mathrm{rad}$.

@@

\subsection{Toy model for realistic thermal relaxation}\label{toy_relax}
To unify $t_\mathrm{cool}$ and $t_\mathrm{diff}$, we define the
effective thermal relaxation timescale in linear theory as
\begin{align}\label{tc_def}
  t_c &\equiv t_\mathrm{cool} + t _\mathrm{diff} =
  \frac{l_\mathrm{rad}^2}{3\eta} + \frac{1}{\eta k_x^2},  
\end{align}
where the relation between $t_\mathrm{cool}$, $l_\mathrm{rad}$ and
$\eta$ can be inferred from Eq. \ref{lrad}, Eq. \ref{tc_newton_cool}
and Eq. \ref{eta_def}. Eq. \ref{tc_def} is a simple prescription so
that for small scales (large $k_x$), $t_c\to t_\mathrm{cool}$, while
for large scales (small $k_x$), $t_c\to t_\mathrm{diff}$.

@@

\subsection{Numerical example}\label{mmsn_example}
Consider  the MMSN at $r=50\mathrm{AU}$. Eq. \ref{mmsn_epsilon} gives
$\epsilon \simeq 
0.067$ which implies, using Eq. \ref{bcrit_mmsn}, a critical 
relaxation timescale of  
\begin{align*}
  \beta_\mathrm{crit} \simeq 0.072. 
\end{align*} 
To estimate a thermal relaxation timescale, we set $\khat=10$ 
in Eq. \ref{beta_mmsn} to obtain 
\begin{align*}
  \beta \simeq 0.049.
\end{align*}
(In fact, Fig. \ref{mmsn_bcrit_bcool} shows that $\beta$  is not
sensitive to $\khat$ at $r=50\mathrm{AU}$ for $\khat\gg1$.)
The disk is unstable to the fundamental isothermal VSI, but not
strongly.

### 2015-01-27 linear_iso.tex
regular update
completed rough summary

\section{Analytical discussion of vertically isothermal disks} 
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $t_c$, or any value of $\gamma$ with
$t_c = 0$) and adiabatic perturbations ($\gamma\neq 1$ and
$t_c\to\infty$), before considering the effect of thermal relaxation.

@@

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to the replacement
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
    &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left(1-\chi\right) 
   \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

>>>>

\section{Analytical discussion of vertically isothermal disks}\label{analytical}
We begin with a discussion of vertically isothermal disks
where $\Gamma=1$, then $c_s$ is a constant in the shearing box. We
examine separately the limiting case of isothermal perturbations
($\gamma\equiv1$ and any value of $t_c$, or any value of $\gamma$ with
$t_c = 0$) and adiabatic perturbations ($\gamma\neq 1$ and
$t_c\to\infty$), before considering the effect of thermal relaxation.

@@

\begin{enumerate}
\item We consider eigenvalues such that
  $|\sigma^2|\ll \kappa^2$. Then we can replace $D=\kappa^2 -\sigma^2\to
  \kappa^2$. % The problem simplifies because the eigenvalue now only
  % appears in the last term in Eq. \ref{iso_ode}.
  We call this the
  low-frequency approximation. 
\item We are mainly interested in the effect of vertical shear, i.e. $q\neq
  0$. This is already captured explicitly in the linearized equations through
  the term $d\Omega/dz$. We therefore simplify
  further by ignoring the vertical dependence of $\Omega$ and $\kappa$
  elsewhere. For clarity, we replace $\kappa^2$ with $\Omega_k^2$. 
  % , although the discussion
  % is unchanged if we replaced 
  % $\kappa^2$ with $\kappa^2(0)$, for example.    
\end{enumerate}
Making both of these approximations is equivalent to setting 
$D\to\Omega_k^2$.  After making this replacement, we eliminate $W$ and
$Q$ between  Eq. \ref{lin_vz} and Eq. \ref{ode_w}---\ref{ode_Q},
making use of Eq. \ref{vertical_shear}, to
obtain an equation for $\delta v_z$ when $\Gamma=1$,
\begin{align}
  0 =& \frac{d^2\delta v_z}{dz^2} + \left(1 + \frac{\ii k_x c_s^2
      q}{\Omega_k^2r}\right)\frac{d\ln\rho}{dz}\frac{d\delta
    v_z}{dz} \notag\\
  &+ \left[\sigma^2\left(\frac{k_x^2}{\Omega_k^2} +
      \frac{\chi}{c_s^2}\right) + \left(\chi + \frac{\ii k_x c_s^2
        q}{\Omega_k^2r}\right)\frac{d^2\ln\rho}{dz^2}\right. \notag\\
    &\phantom{+=}\left.-
    \frac{c_s^2}{\Omega_k^2}\left(\frac{d\ln\rho}{dz}\right)^2\left(k_x^2 -
      \frac{\ii k_x q}{r}\right)
   \left(1-\chi\right) 
   \right]\delta v_z,\label{vertiso_gov}
\end{align}
where
\begin{align}
  \chi \equiv \frac{1-\ii\sigma t_c}{1-\ii\sigma t_c\gamma}.
\end{align}

### 2015-01-27 summary.tex
completed rough summary

For non-isothermal ($\gamma\neq1$) perturbations with intermediate
thermal relaxation timescales, the general dispersion relation
(Eq. \ref{relax_disp}) is complicated and requires a numerical
study. However, for the fundamental VSI, characterized by  $\delta
v_z\sim\mathrm{constant}$ near $z=0$, some simplification is
possible. In particular, we quantified the thermal relaxation
timescale $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below
which small-scale perturbations ($\khat\gg 1$) may be regarded as isothermal
and the standard VSI operates. We believe this explicit expression for
$\beta_\mathrm{crit}$ is the most useful result from our
investigation.

%summary of results

@@

\subsection{Caveats and outlooks}

>>>>

For non-isothermal ($\gamma\neq1$) perturbations with intermediate
$\beta$, the general dispersion relation
(Eq. \ref{relax_disp}) is complicated and requires a numerical
treatment. However, for the fundamental VSI, characterized by  $\delta
v_z\sim\mathrm{constant}$ near $z=0$, some simplification is
possible. In particular, we quantified the thermal relaxation
timescale $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below
which small-scale perturbations ($\khat\gg 1$) may be regarded as isothermal
and the standard VSI operates. We believe this explicit expression for
$\beta_\mathrm{crit}$ is the most useful result from our
investigation.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. Numerical solutions are mostly consistent
with our analytical expectations, except for perturbations with small
radial wavenumber ($\khat\to0$) where the low-frequency approximation
is inappropriate.

We briefly studied vertically non-isothermal disks using our linear
code. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI operates and is insensitive to thermal relaxation. When a
stable stratification is introduced, the dependence of VSI growth
rates on the thermal relaxation timescale is qualitatively similar to
that observed for vertically isothermal disks.

@@

We applied our linear results to assess the relevance of the VSI in
protoplanetary disks, specifically the Minimum Mass Solar Nebulae (MMSN),
using disk models described in \cite{chiang10}. We find the MMSN is
unstable to the VSI at $10$s of AU, with characteristic
growth times of $10$s of orbits. However, we note the strong
temperature dependence of the physical thermal relaxation timescale
in the disk model ($\beta$ decreases as $T^{-5}$ for perturbations with small
radial scale). Thus, increasing the disk temperature can signifcantly 
shorten the typical growth times for the VSI. This is also attributed
to a larger value of $\beta_\mathrm{crit}$, since increasing $T$ also
increases the aspect-ratio $\epsilon$.

\subsection{Future extensions} 
We highlight below some improvements to the present  
linear models that should be pursued:

\emph{Realistic thermal evolution.} We have treated the thermal
relaxation time $t_c$ as a constant for simplicity. However, a
physical expression for $t_c$ will depend on 
$z$. For example, the Newtonian cooling timescale
(\S\ref{newton_cool}) will depend on $z$ for vertically non-isothermal
disks, and through the opacity since 
$\kappa_d=\kappa_d(\rho,T)$ in general.

For vertically isothermal disks and a temperature-dependent opacity
model, as considered in \S\ref{toy_relax} and Appendix \ref{mmsn},
there is still a vertical dependence of the effective thermal relaxation time
because the radiative diffusion coefficient depends on $\rho$. We
artificially discarded this vertical dependence by 
replacing $\rho\to\Sigma/H_\mathrm{iso}$, but can restore it
by making the transformation
$\beta(\khat)\to\beta(\khat/\sqrt{2\pi}\hat{\rho})\equiv
\tilde{\beta}(\khat,z)$ in  Eq. \ref{real_beta}, where
$\hat{\rho}=\rho/\rho(z=0)$.  We have repeated the example calculation
in \S\ref{mmsn_example}  with $\beta\to\tilde{\beta}$, but found
similar results.

A proper treatment for radiative diffusion will ultimately involve
the Laplacian operator. In this case additional boundary conditions 
must be supplemented. We will consider this in a future work.

\emph{Additional physics.} Depending on where the VSI operates in a
disk, additional effects may become important. For example, in the
outer parts of a protoplanetary disk, where we estimate the VSI may
operate, disk self-gravity (SG) may play a role. Specifically, the vertical
shear profile depends on the vertical density stratification
(Eq. \ref{vertical_shear}), which wil be enhanced by vertical SG,
which will enhance the VSI. Inclusion of the Poisson equation should
therefore be considered.

We have neglected viscosity. \cite{nelson13} have shown that an
alpha viscosity of $\lesssim O(10^{-4})$ is required for the VSI to
operate. It is of interest to quantify the effect of viscosity
on the linear VSI. This would help identify, more accurately, the
appropriate regions in a protoplanetary disk that are sufficiently
laminar to develop the VSI.  This problem is, however, complicated by
the fact that proper inclusion of viscosity introduces viscous heating
in the energy equation.

% \emph{Theoretical generalizations.} Our analytical discussion
% (\S\ref{analytical}) can be extended in several ways. One could
% relax the low-frequency approximation to improve the agreement between
% numerical and analytical eigenfrequencies at small $\khat$. However,
% such modes are radially more global, and would need a global
% cylindrical disk model to study them. Another minor improvement is to

%analytical extensions -> nonisothermal, all frequency, non-keplerian,
%diffusion 
%viscosity self-gravity 
%disk models: polytropic connected with isothermal
